---
title: "Statistical inference with the GSS data"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
library(statsr)
```

### Load data

Make sure your data and R Markdown files are in the same directory. When loaded
your data file will be called `gss`. Delete this note when before you submit 
your work. 

```{r load-data}
load("gss.Rdata")
```



* * *

## Part 1: Data

**Data:** The data was collected in rural and urban areas in the US, in the form
of a survey/questionare From 1972-2018. There are a total of 64,814 completed 
interviews.  Each survey was an independently drawn sample of English-speaking 
persons 18 years of age or over, living in non-institutional arrangements within 
the United States. Spanish-speakers were also added to the target population. 
Block quota sampling was used and full probability sampling was employed. The 
data from the interviews were processed according to standard NORC procedures.

**Case:** Here in the data, the cases in the people residing in the US who are 
interviewed in their household.

**Variables:** Two variables `race` and `partyid`. `race` represents the races 
of respondent. Data type of `race` is numeric. `partyid` represents political 
party affiliation. Data type of `partyid` is also numeric. 

**Type of study:** GSS is a questionare/survey done by interviewing Americans 
that were random sampled from the US adult population. So, it is an 
observational study. 

**Scope of inference - generalizability:** The adult people of the US.

**Scope of inference - causality:** Since this is an observational study, there
is no causality.

* * *

## Part 2: Research question

**Research Question:** Does the data suggest that there is an association 
between the variable `race` and `partyid`? This question explores is there a 
relationship between race and political party affiliation. This research 
question will shed light into the racial support that each party has. 

* * *

## Part 3: Exploratory data analysis

NOTE: Insert code chunks as needed by clicking on the "Insert a new code chunk" 
button (green button with orange arrow) above. Make sure that your code is visible
in the project you submit. Delete this note when before you submit your work.


```{r view_data_race}
summary(gss$race)
```
The data showcases 46350 white persons and 7926 black persons. 

```{r view_data_partyid}
summary(gss$partyid)
```
There are 327 NA in the data. 

**Contingency Table:** Relationship between the 2 categorical variable
```{r table_race_partyid}
table(gss$partyid, gss$race)
```

**Contingency Table with Proportion:**  For categorical data, it is appropriate 
to use percentage(proportions) instead of counts to see the relationship.
```{r prop_table_race_partyid}
race_partyid = table(gss$race,gss$partyid)

partyid_race = table(gss$partyid,gss$race)

prop.table(race_partyid)
```

**Visualization:** Using bar_plot and mosaic_plot can help understand the data
better and understand the relationship between each category.
```{r barplot_partyid}
barplot(table(gss$partyid), beside=TRUE, xlab="Political Party Affiliation", ylab="Count", main="Bar Plot of Political Party Affiliation in the US")
```

```{r barplot_race_partyid}
barplot(race_partyid, beside=TRUE, legend.text=TRUE, xlab="Political Party Affiliation", ylab="Racial Status", main="Bar plot of racial status and political party affiliation")
```

```{r mosaicplot_race_partyid}
mosaicplot(partyid_race, shade = TRUE)
```

* * *

## Part 4: Inference

NOTE: Insert code chunks as needed by clicking on the "Insert a new code chunk" 
button (green button with orange arrow) above. Make sure that your code is visible
in the project you submit. Delete this note when before you submit your work.

**Hypothesis**
$$H_0:$$ Race and Political Party Affiliation are independent. 
$$H_1:$$ Race and Political Party Affiliation are dependent.
We choose the 5% significance level.
$$\alpha = 0.05$$

**Conditions**
Independence: Random sampling or Random assignment, If sampling without 
replacement, n<10% of population, Each case only contributes to one cell in the 
table Sample Size / Skew: Each particular scenario (i.e., cell) must have at 
least 5 expected cases.

**Inferential Method**
We use statistical inference, as the inference conditions are met for hypothesis
testing (Random sampling and large sample size). As the 2 variables are 
categorical and one has more than 2 levels and we are testing for independence, 
we will be using Chi-square Independence Test for which the conditions are met. 

```{r}
check_expected_count = chisq.test(partyid_race)
check_expected_count$expected
```
Simulation: We check the expected value for each cell assuming that the null
hypothesis is true. 

**Perform Inference**
For test statistic, Chi.square independence test statistic is calculated to 
decide on either to reject the null hypothesis or fail to reject the null 
hypothesis. 
```{r}
chisq.test(partyid_race)
```

Since p_value < 2.2e-16, which is extremely small and smaller than 0.05, we reject
the null hypothesis. Since the null hypothesis is rejected, it implies that 
there is a dependence/association between race and party affiliation. 

**Different Methods**
Simulation: 
```{r}
chisq.test(partyid_race, simulate.p.value=TRUE, B=10000)$p.value
```
The p_value is extremely small and is smaller than the significance level. 
Therefore, we still have enough evidence to reject the null hypothesis.

* * * *

## Part 5: Conclusion
Summary: As per hypothesis test results, the null hypothesis was rejected. 
Therefore, there is a dependency or association between `race` and `partyid`. 